Variance reduction for MCMC methods via martingale representations

### 2020-03-04 aap/Revision/vr-mcmc-revision-new.tex
experiments description

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\) and \(f(x) = \sum_{i=1}^d x^2_i\) we have used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent "training" trajectories of length $n = 10^4$.  Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.

@@

In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) = \mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{-\theta^T \mathsf{X}_i})} + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.

>>>>

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
\]
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.  Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, since they are used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.

@@

In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) = -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2, \quad
\nabla U(\theta) = -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials for the same reasons as in the Gaussian mixtures example.
